Mixture Models for Single Cell Assays

### 2012-03-05 MIMOSA.tex
minor updates
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\subsection{Simulations}
We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations. Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from the HVTN054 data set, we simulated data from the constrained model with 500 observations, a response rate of 40\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$. The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test. (Figure~\ref{fig:simulations}). The nominal vs observed false discovery rate was also examined for the models and for Fisher's exact test, as well as the accuracy of the estimated prior distributions. In order to assess whether the unconstrained model could be used to effectively fit data from a constrained model and benefit from faster computations using exact closed form expressions rather than Monte-Carlo integration, we fit the same data simulated from the constrained model using the unconstrained model (Figure~\ref{fig:simulations}.

@@

\section*{Results}
\subsection*{Simulations}
We ran simulation studies to assess the performance of the constrained and unconstrained models, as described in the Methods. We found that the both the constrained and unconstrained MIMOSA models out--performed Fisher's exact test with respect to sensitivity and specificity at all values of $N$, and that the false discovery rate observed for the mixture model more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations}). Furthermore, both models gave reasonable estimates of the true hyper--parameters (Figure~\ref{fig:simulations}).

@@

\subsection*{HVTN054 ICS data}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{Figures/simulations_all} 
   \caption{Performance of the constrained and unconstrained Beta--binomial mixture model vs Fisher's exact test on simulated data. Data were simulated from a constrained model with hyper--parameters estimated from a real data of Gag1 stimulated, CD4+, IL2 expressing T--cells on day 28 from the HVTN054 trial. For total cell counts from 10,000 to 150,000, we simulated ten data sets each of 500 observations with a response rate of 40\%. The performance, measured by the AUC (area under the curve), of the constrained and unconstrained Beta--binomial mixture model compared to Fisher's exact test is shown in the first panel, as a function of increasing number of cells. The estimated and true prior distributions are shown in the second panel for one simulated data set, with N=150,000 events. The ROC curve for Fisher's exact test and the constrained and unconstrained Beta--binomial model for the same simulation are shown in the third panel. The observed vs expected false discovery rate for Fisher's exact test and the constrained and unconstrained Beta--binomial model are shown in the fourth panel.}
   \label{fig:simulations}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{Figures/simulations_violatedmodel} 
   \caption{Sensitivity to departures from model assumptions. We generated data from variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions. The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta--prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity). The AUC for the unconstrained MIMOSA model and Fisher's exact test as a function of event count are shown in the first panel. The estimated and true prior distributions are shown in the second panel. The ROC curves for one data set and the observed and true false discovery rates for the model and Fisher's exact test are shown in the third and fourth panels.}
   \label{fig:simulations_trunc}
\end{figure}

\subsection*{HVTN054 ICS Data}

>>>>

\subsection*{Simulation Studies}
We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations. Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from the HVTN054 data set, we simulated data from the constrained model with 500 observations, a response rate of 40\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$. The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test. (Figure~\ref{fig:simulations}). The nominal vs observed false discovery rate was also examined for the models and for Fisher's exact test, as well as the accuracy of the estimated prior distributions. In order to assess whether the unconstrained model could be used to effectively fit data from a constrained model and benefit from faster computations using exact closed form expressions rather than Monte-Carlo integration, we fit the same data simulated from the constrained model using the unconstrained model (Figure~\ref{fig:simulations}.

@@

\section*{Results}
\subsection*{Simulations Show MIMOSA Outperforms Fisher's Exact Test Event When Model Assumptions are Violated}
We ran simulation studies to assess the performance of the constrained and unconstrained models, as described in the Methods. We found that the both the constrained and unconstrained MIMOSA models out--performed Fisher's exact test with respect to sensitivity and specificity at all values of $N$, and that the false discovery rate observed for the mixture model more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations}). Furthermore, both models gave reasonable estimates of the true hyper--parameters (Figure~\ref{fig:simulations}).

@@

\subsection*{MIMOSA Outperforms Fisher's Exact Test in Real--World Vaccine Trial Data from HVTN054}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{Figures/simulations_all} 
   \caption{Performance of the constrained and unconstrained Beta--binomial mixture model vs Fisher's exact test on simulated data. Data were simulated from a constrained model with hyper--parameters estimated from a real data set of Gag1 stimulated, CD4+, IL2 expressing T--cells on day 28 from the HVTN054 trial. For total cell counts from 10,000 to 150,000, we simulated ten data sets each of 500 observations with a response rate of 40\%. The performance, measured by the AUC (area under the curve), of the constrained and unconstrained Beta--binomial mixture model compared to Fisher's exact test is shown in the first panel, as a function of increasing number of cells. The beta distributions for the estimated and true hyper parameters are shown in the second panel for one simulated data set, with N=150,000 events. The ROC curve for Fisher's exact test and the constrained and unconstrained Beta--binomial model for the same simulation are shown in the third panel. The observed vs expected false discovery rate for Fisher's exact test and the constrained and unconstrained Beta--binomial model are shown in the fourth panel.}
   \label{fig:simulations}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{Figures/simulations_violatedmodel} 
   \caption{Sensitivity to departures from model assumptions. We generated data from variant of the constrained model ($p_s>p_u$) where the proportions were simulated from truncated normal distributions on $(0,1)$, rather than from beta distributions. The mean and variance of the normal distributions was given by $\mu=\alpha/(\alpha+beta), \sigma^2=(\alpha\beta)/((\alpha+beta)^2(\alpha+\beta+1))$, where $\alpha,\beta$ are the Beta--prior hyper parameters for the MIMOSA model estimated from real data ($s, u$ subscripts omitted for brevity). The AUC for the unconstrained MIMOSA model and Fisher's exact test as a function of event count are shown in the first panel. The beta distributions for the estimated and true hyper parameters  are shown in the second panel. The ROC curves for one data set and the observed and true false discovery rates for the model and Fisher's exact test are shown in the third and fourth panels.}
   \label{fig:simulations_trunc}
\end{figure}
